Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-08-11 paper/paper.tex
new refs (pomdp) + new ideas (pomdp, hippocampal stuff, etc.)

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Neurological patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological relevance assessment
\citep{amygdala_bzdok}.

@@

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensory perception.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

\begin{wrapfigure}{r}{.3\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

@@

\subsection{Markov decision processes}
Model-free RL has had great success in many real-world problems,
including robotics \citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015},
and strategic board games like the breakthrough results upon recently on the game of Go \citep{silver2016mastering}, considered to be a golden benchmark problem in artificial intelligence.
We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in active learning.

@@

@@

In the human brain,
the HC could contribute to synthesizing imagined sequences of world states,
actions and rewards \citep{aronov2017, chao2017interaction, boyer2008evolutionary}.
These simulations of experience batches
would be used to update the value function, without ever looking inside the black box describing the model's dynamics \citep{lavilleon2015}.
This would be a simple control algorithm by evaluating all legal actions and selecting the action with
highest expected cumulative rewards.
In MDPs, MCMC simulation provides an effective mechanism both for tree search and for belief-based state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible \citep{silver2016mastering}.
% Much recent work points to MCMC or stochastic sampling-based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. \textbf{XXX elvis: refs needed in last sentence. Plus, should be really bring up Bayesian?}

>>>>

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Neurological patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological relevance assessment
\citep{Averbeck2017,amygdala_bzdok}.

@@

\begin{wrapfigure}{r}{.35\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensory perception.
\citep{starkweather2017dopamine,o2006making,o2006making}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\subsection{Interlude: Rethinking the Action-State-Reward model of RL}
``Remarkably, Barto proposes to decouple extrinsic rewards (i.e generated by the external environment) from intrinsic reward (generated from withing the "organism")
biologically, extrinsic rewards would correspond to so-called "sensations"
while intrinsic rewards would correspond to "reward signals".

Thus there are two enviroments: the external environment, and the internal environment built-into the "organism". The organism takes highly level DECISIONS (the DMN can confortably sit here!) which act on the internal environment (this environment probably corresponds to low-level subsystems like motor, etc.) which then transform theses  into ACTIONS executed in the external environment. The external environment generates sensations which get mapped into reward signals (via the internal environment), etc.
Il y a 6 min
Guillaume Dumas (deep.introspection@gmail.com)
Yeah, good for creating Skynet :D!
(Half joking 🙃)
Guillaume • Il y a 6 min
haha
The interesting thing is that with this rethinking, the organism can have a perception-action loop even in the absence of external stimulus.
This is called curiosity.''

\subsection{Markov decision processes}
Model-free RL has had great success in many real-world problems,
including robotics \citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015},
and strategic board games like the breakthrough results upon recently on the game of Go \citep{silver2016mastering}, considered to be a golden benchmark problem in artificial intelligence.
In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in active learning.

@@

\begin{mdframed}
\paragraph*{Why MDPS ?} At this point, it is righteous to question the applicability of MPDs as a model for something as complex as human behavior ?
Indeed a critique could argue that Markov chains are state-machines without memory. These may describe insects, or other behaviorally primitive beings like ATM machines, insects, etc., but won't be sufficient to describe human beings, as for example, human behavior is heavily influenced not only by the current state but by past states as well (memory).

\paragraph*{Answer:} We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.

Rational probabilistic planning can be reformulated
as a standard memory Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.
This adds the capacity for memory, since then, the next state depends not only on the current state but also on a finite number of past states.

Also, it should be noted that MDPs have been successfully used as neural model for animal behavior. One can mention ~\citep{Rao2010}, just to name a few.
\end{mdframed}

@@

In the human brain,
the HC could contribute to synthesizing imagined sequences of world states,
actions and rewards \citep{aronov2017, chao2017interaction, boyer2008evolutionary}.
These simulations of experience batches
would be used to update the value function, without ever looking inside the black box describing the model's dynamics \citep{lavilleon2015}.
This would be a simple control algorithm by evaluating all legal actions and selecting the action with
highest expected cumulative rewards.As a result, quick-to-learn instance-based control policies serve as a rough approximation while a slower more generalisable decision-making system system is trained up~\citep{pritzel2017neural,lengyel2008hippocampal}.
In MDPs, MCMC simulation provides an effective mechanism both for tree search and for belief-based state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible \citep{silver2016mastering}.
% Much recent work points to MCMC or stochastic sampling-based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. \textbf{XXX elvis: refs needed in last sentence. Plus, should be really bring up Bayesian?}
